---
title: "From experiment to trustworthy comparison"
subtitle: "BIOSCI 504 · Day 2"
author: "Denis O'Meally"
institute: "City of Hope"
format:
  revealjs:
    theme:
      - default
      - brand
    slide-number: c/t
    transition: none
    controls: true
    progress: true
    hash: true
    history: true
    center: false
    width: 1600
    height: 900
    margin: 0.08
    code-overflow: wrap
    embed-resources: true
    footer: "BIOSCI 504 · From experiment to trustworthy comparison"
execute:
  warning: false
  message: false
---

## By the end

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
library(tidyverse)

data(mouse_trial, package = "BIOSCI504")
data(course_data_dictionary, package = "BIOSCI504")

illustrative_comparison <- tibble(
  comparison_group = rep(c("A", "B"), each = 5),
  outcome = c(1, 2, 3, 4, 5, 2.8, 2.9, 3, 3.1, 3.2)
)

illustrative_outcome_plot <- ggplot(
  illustrative_comparison,
  aes(x = comparison_group, y = outcome, color = comparison_group)
) +
  scale_color_manual(values = c(A = "#177BFB", B = "#0757A8")) +
  labs(
    x = "Illustrative group",
    y = "Illustrative outcome",
    caption = "Illustration only — not mouse_trial results"
  ) +
  theme_minimal(base_size = 24)
```

You should be able to:

1. state what one row represents;
2. inspect a table before changing it;
3. choose an operation that matches the biological question;
4. inspect a derived variable before comparing it;
5. distinguish a result, its interpretation and the evidence supporting it.

::: {.notes}
Lecture 3 prepares the reasoning needed for the tabular-comparison exercise. It
does not provide the completed treatment comparison. Canonical table outputs
were built from `BIOSCI504` commit
`3730f8b813668f317689ac56fabb299c9a1b072e`.

The sequence adapts current Carpentries teaching moves: predict before reveal,
identify the observational unit, inspect with a purpose, construct a
transformation visibly, inspect the new state, and retain individual
observations beside summaries.

These are reasoning outcomes. Syntax supports them but is not the main target.

Ask: If R executes every line without an error, which of these outcomes has
already been established? Answer: none.
:::

## A recovery study

```text
48 mice
   │
   ├── baseline weight
   │
   ├── control or nutritional treatment
   │
   └── final weight

sex · cage · experimental batch
```

Did treated mice gain more weight than control mice?

Treatment varies within cages; treatment and cage are not the same grouping.

::: {.notes}
Describe the experiment before showing the table. The study includes biological
units, an assigned treatment, paired measurements and context that may
contribute to variation.

Treatment and control mice occur within cages. That structure is why cage is
experimental context rather than a substitute label for treatment.

Do not discuss the observed treatment result. The question is the starting
point, not a conclusion.
:::

## Before seeing the table

Predict:

- What should one row represent?
- Which columns are needed?
- Which column should identify a mouse?
- Where should the unit of weight be recorded?

::: {.notes}
Take answers before revealing the data. This follows the Carpentries practice
of connecting a data representation to its domain before introducing
operations.

Expected ideas include mouse ID, treatment, baseline and final weight. Sex,
cage and batch may emerge as useful experimental context.
:::

## Four rows from the real table {.smaller}

```{r}
#| echo: false

mouse_trial |>
  slice_head(n = 4) |>
  select(
    mouse_id,
    treatment,
    sex,
    baseline_weight_g,
    final_weight_g,
    cage,
    batch
  ) |>
  knitr::kable()
```

::: {.notes}
This is a genuine sample from `mouse_trial`, not a mock-up. Ask learners to
locate the identifier, grouping variable, measurements and experimental
context.

A preview establishes that these values exist. It does not establish the size,
completeness, uniqueness or plausible range of the whole table.
:::

## What does one row represent?

Choose one:

1. one weight measurement;
2. one mouse;
3. one cage;
4. one treatment group.

::: {.notes}
Answer: one mouse. Baseline and final measurements occupy different columns on
the same mouse row.

This question determines how identifiers, duplicates, sample size and within-
mouse change should be interpreted. Return to it throughout the lecture.
:::

## Rows, columns and cells

| Part | Meaning in this table |
|---|---|
| row | one observed mouse |
| column | one recorded or derived variable |
| cell | one value for one mouse and variable |
| `mouse_id` | the identifier for the observational unit |

One value per cell; units belong in metadata or the column name.

::: {.notes}
Adapted from Data Carpentry's spreadsheet-ecology tidy-table rules: columns are
variables, rows are observations and cells contain values.

Source: https://datacarpentry.github.io/spreadsheet-ecology-lesson/01-format-data.html
:::

## Variables have different roles

| Role | Variables |
|---|---|
| identifier | `mouse_id` |
| assigned group | `treatment` |
| biological/context variables | `sex`, `cage`, `batch` |
| continuous measurements | `baseline_weight_g`, `final_weight_g` |

The role follows from the experiment, not merely the R storage type.

::: {.notes}
Connect to Lecture 2: character and double describe storage. Identifier,
treatment, context and measurement describe scientific roles.

`treatment`, `sex`, `cage` and `batch` are categorical variables even though
their values are stored as character strings. Weight is continuous.
:::

## The dictionary is part of the data {.smaller}

```{r}
#| echo: false

course_data_dictionary |>
  filter(dataset == "mouse_trial") |>
  transmute(
    variable,
    description,
    units = replace_na(units, "—"),
    allowed_values = replace_na(allowed_values, "—")
  ) |>
  knitr::kable()
```

::: {.notes}
A column name is not a complete definition. The dictionary records intended
meaning, units and permitted categories.

Ask: Which source would you use to decide whether `27500` is a plausible final
weight—the R type or the dictionary? The type can only say it is a double; the
dictionary supplies grams and biological meaning.
:::

## Checkpoint: read the representation

For the cell in row 2 under `final_weight_g`:

- What does its row represent?
- What does its column represent?
- What does the value represent?
- Where does its unit come from?

::: {.notes}
Assess the whole room. One row is mouse M002; the column is post-treatment
weight; the cell is that mouse's recorded final weight; the `_g` suffix and
dictionary establish grams.

Do not move on until row versus measurement is clear.
:::

## A preview is not an inspection

```r
head(mouse_trial)
```

can show examples of values.

It cannot establish:

- the total number of records;
- whether every ID is unique;
- missingness elsewhere;
- complete group counts;
- the full range of a measurement.

::: {.notes}
`head()` is useful, but only samples the first rows. This is why inspection
needs checks over the complete object.

Source: https://datacarpentry.github.io/R-ecology-lesson/instructor/how-r-thinks-about-data.html
:::

## Always inspect

| Check | A failure it could reveal |
|---|---|
| dimensions | missing or extra records |
| names and types | wrong schema or import |
| distinct IDs | duplicated observational units |
| missing values | incomplete measurements |
| group counts | imbalance or miscoding |
| ranges and units | impossible or mis-scaled values |

Inspection is a set of questions, not one command.

::: {.notes}
This is the standing rule for the session. Narrate the purpose of each check.
A long list of commands without expected behavior is not verification.

Ask learners to name a result that would concern them before showing code.
:::

## Inspect structure

```r
dim(mouse_trial)
names(mouse_trial)
glimpse(mouse_trial)
```

These answer different questions:

- How large is the table?
- Does it have the expected schema?
- How are the variables represented?

::: {.notes}
Live-code these in RStudio. The canonical table has 48 rows and seven columns,
but let the live output provide those values.

`glimpse()` is the tidyverse counterpart to the `str()` used in the
Carpentries source. It shows structure and sample values, not every data-quality
condition.
:::

## Check the observational unit

```r
mouse_trial |>
  summarise(
    rows = n(),
    distinct_mice = n_distinct(mouse_id)
  )
```

What relationship do you expect between the two results?

::: {.notes}
If one row represents one mouse, rows and distinct IDs should match. A mismatch
does not tell you the cause; it tells you the representation needs
investigation.

Ask for the expected relationship before execution. Do not introduce the
duplicate challenge fixture.
:::

## Check groups and missingness

```r
mouse_trial |>
  count(treatment)

mouse_trial |>
  summarise(
    across(
      c(baseline_weight_g, final_weight_g),
      ~ sum(is.na(.x))
    )
  )
```

::: {.notes}
Live-code or reveal the commands, but do not make learners copy every output
into the slides before the exercise.

Group counts establish the denominator. Missingness should also be checked by
scientifically relevant groups later; an overall total can conceal a group-
dependent problem.
:::

## Check ranges against units

```r
mouse_trial |>
  summarise(
    across(
      ends_with("_weight_g"),
      ~ list(range(.x, na.rm = TRUE))
    )
  )
```

The suffix `_g` makes the expected scale testable.

::: {.notes}
The exact command is less important than comparing observed ranges with the
dictionary and biological expectation.

An R value can be a perfectly valid double and still use the wrong unit. This
is one form of wrong-but-running data.
:::

## Checkpoint: choose the revealing check

Which check is most direct?

| Concern | Candidate check |
|---|---|
| the same mouse appears twice | ? |
| one weight uses another unit | ? |
| final weights are missing mainly in one group | ? |

::: {.notes}
Expected answers: compare rows with distinct IDs or inspect duplicates; inspect
ranges/plots against documented units; count missingness within treatment.

More than one check may contribute. Ask learners to explain what evidence each
one would provide.
:::

## The question determines the operation

:::: {.columns}

::: {.column width="48%"}
### Final weight

Are treated mice heavier at the end?
:::

::: {.column width="48%"}
### Weight change

Did treated mice gain more weight?
:::

::::

These are different biological questions.

::: {.notes}
This is the conceptual centrepiece. Final weight contains starting weight plus
change during the study. Weight gain isolates the within-mouse difference.

Neither question is inherently invalid. The operation must match the question
that was specified.
:::

## Two analytical targets

```text
final weight             final weight − baseline weight
     │                                  │
post-treatment level                  change during study
```

A mouse can finish heavier because it started heavier, gained more, or both.

::: {.notes}
Use a verbal example: two mice can gain the same amount and finish at different
weights because they started at different weights.

Do not show the treatment-specific result. The point is the target of the
comparison.
:::

## Valid code can answer the wrong question

Question: **Did treated mice gain more weight?**

```r
mouse_trial |>
  group_by(treatment) |>
  summarise(
    n = sum(!is.na(final_weight_g)),
    mean_final_g = mean(final_weight_g, na.rm = TRUE)
  )
```

This code runs. Which question does it actually answer?

::: {.notes}
It compares final weight, not gain. This is the required wrong-but-running
example: there is no syntax error, and the output can look reasonable, but it
does not implement the stated analytical target.

Ask learners to name the mismatch before correcting it. Do not run or reveal
the canonical group means.
:::

## Name the quantity before coding

For each mouse:

$$
\text{weight change (g)} = \text{final weight (g)} - \text{baseline weight (g)}
$$

Expected sign: positive for gain, negative for loss.

::: {.notes}
Start in words, then arithmetic, then an R variable name. Units remain grams
because grams are subtracted from grams.

Ask for the expected sign before showing the transformation. This gives the
subsequent range check a scientific expectation.
:::

## A derived variable is a transformation

```r
mouse_trial |>
  mutate(
    weight_change_g = final_weight_g - baseline_weight_g
  )
```

`mutate()` returns a new table state with a new column.

::: {.notes}
Live-code this without assigning it first. Let the intermediate table print so
the transformation is visible, following the current R Ecology teaching
sequence.

Source: https://datacarpentry.github.io/R-ecology-lesson/instructor/working-with-data.html#making-new-columns-with-mutate
:::

## Inspect the new state

```r
mouse_trial |>
  mutate(
    weight_change_g = final_weight_g - baseline_weight_g
  ) |>
  summarise(
    type = typeof(weight_change_g),
    missing = sum(is.na(weight_change_g)),
    minimum = min(weight_change_g, na.rm = TRUE),
    maximum = max(weight_change_g, na.rm = TRUE)
  )
```

::: {.notes}
Build the pipeline one step at a time. Ask what each check could reveal: type
confirms representation, missingness confirms propagation, and range checks the
sign and scale against the biological expectation.

Stop before grouping by treatment or producing the exercise comparison.
:::

## Missing inputs propagate

```text
final weight     baseline weight     change
    25.3 g    −       NA          =    NA
      NA      −      24.2 g       =    NA
```

The mouse remains a row. The derived measurement is missing.

::: {.notes}
Connect directly to Lecture 2. Arithmetic cannot reconstruct an unobserved
input. The derived column therefore has missing values wherever either source
measurement is missing.

Do not silently delete these rows. Record how missingness changes the effective
sample and whether it differs by treatment.
:::

## Checkpoint: predict the transformation

For one mouse:

```text
baseline = 24.1 g
final    = 25.3 g
```

Before calculating:

- What sign should the change have?
- What is its value and unit?
- What does the row represent after adding the column?

::: {.notes}
Answer: positive, 1.2 g, and the row still represents the same mouse. Adding a
derived variable changes the columns, not the observational unit.

This is a small-case check against the transformation before it is scaled to
the whole table.
:::

## Plot individual observations first

```{r}
#| echo: false
#| fig-height: 4.8
#| fig-alt: "Two dot plots with five observations each. Group A spans outcomes 1 to 5, while group B is tightly clustered around 3."

illustrative_outcome_plot +
  geom_jitter(width = 0.08, size = 5, show.legend = FALSE)
```

::: {.notes}
This deliberately neutral illustration avoids revealing the exercise result.
Both groups have the same mean, but their spread is very different.

Ask learners what they can see before introducing the mean: sample size,
individual values, spread and an unusual or extreme observation.
:::

## A summary deliberately removes detail

```{r}
#| echo: false
#| fig-height: 4.8
#| fig-alt: "The same two dot plots with diamond mean markers. Both means equal 3 despite the much wider spread in group A."

illustrative_outcome_plot +
  geom_jitter(width = 0.08, size = 5, alpha = 0.45, show.legend = FALSE) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 8,
    show.legend = FALSE
  ) +
  labs(
    caption = "Diamonds show group means; illustration only"
  )
```

Same mean; different evidence.

::: {.notes}
The mean retains one aspect of location and discards the individual pattern.
It should be presented beside, not instead of, the observations.

Ask what would be lost if only the two diamonds were shown.
:::

## Grouped summaries change the table

```r
illustrative_comparison |>
  group_by(comparison_group) |>
  summarise(
    n = n(),
    mean = mean(outcome),
    .groups = "drop"
  )
```

Before running: how many rows should the result contain?

::: {.notes}
Answer: two rows, one per group. The observational unit of this summary table is
now the group, not the individual.

Keep `n` beside the mean so the denominator remains visible. The same logic will
apply in the exercise, but do not run the canonical treatment summary here.

Source: https://datacarpentry.github.io/R-ecology-lesson/instructor/working-with-data.html#the-split-apply-combine-approach
:::

## Uncertainty belongs with an estimate

| Quantity | What it reports |
|---|---|
| observed difference | size and direction in these data |
| confidence interval | effect sizes compatible with the data and model under the stated procedure |
| p-value | how unusual the result would be if a specified null model were true |

None establishes that the design, data handling or code are correct.

::: {.notes}
Keep this statistical treatment modest. A confidence interval describes
uncertainty around an estimate under assumptions; it is not the range containing
95% of individual mice. A p-value is not the probability that the null
hypothesis is true and is not a measure of biological importance.

The exercise should report a difference and appropriate uncertainty, but this
lecture does not calculate its final values.
:::

## Before choosing a test, record the assumptions

- What is the observational unit?
- Are observations independent enough for the proposed comparison?
- How were treatment and measurements assigned or collected?
- Are missing values plausibly ignorable?
- Do spread, shape or unusual values challenge the summary?
- Are all measurements expressed in the documented unit?

::: {.notes}
Mice share cages, so treating every row as wholly independent may be an
approximation. Batch and cage can also contribute variation. State that without
turning this into a mixed-model lecture.

The test choice follows the question, design and observed data. It does not
replace inspection.
:::

## Visualization is verification evidence

A plot can reveal:

- spread hidden by a mean;
- an observation on the wrong scale;
- a group with few usable values;
- an unexpected relationship between baseline and final weight;
- a transformation whose sign or range is implausible.

The figure should be checked before it is interpreted.

::: {.notes}
The plot is not decoration added after the statistics. It is both an analytical
result and a diagnostic artifact.

In the exercise, students should show individual observations and a restrained
summary, then check that the figure agrees with their numerical result.
:::

## Checkpoint: what does the summary row represent?

After `group_by(treatment)` and `summarise()`:

1. Does one row still represent one mouse?
2. What does `n` establish?
3. Which feature of the individual data can the mean conceal?
4. Which assumption could shared cages challenge?

::: {.notes}
Answers: one row represents one treatment group; `n` gives the number of rows or
usable values as defined; a mean can conceal spread, overlap, clusters or
outliers; shared cages can challenge simple independence.

Use the answers to decide whether the class is ready for the exercise.
:::

## Result is not interpretation

:::: {.columns}

::: {.column width="48%"}
### Result

- observed difference
- uncertainty
- sample size
- visual pattern
:::

::: {.column width="48%"}
### Interpretation

- relation to the biological question
- plausible explanations
- design limits
- remaining uncertainty
:::

::::

::: {.notes}
Write the numerical and visual result before supplying a biological story. This
makes it easier to see where observation ends and explanation begins.

Avoid causal language beyond what the treatment assignment and experimental
design support.
:::

## The design limits the claim

The table records:

- assigned treatment;
- paired weights;
- sex, cage and batch;
- missing measurements.

It does not automatically establish:

- the cause of every individual difference;
- independence within cages;
- why a measurement is missing;
- generality beyond the studied mice and conditions.

::: {.notes}
The limits are properties of the design and recorded evidence, not a ritual
disclaimer. Ask which additional data or design feature would change a
particular conclusion.
:::

## Why I trust this

| Evidence | What it supports |
|---|---|
| rows match distinct mouse IDs | observational unit represented once |
| missingness recorded by group | effective sample is visible |
| ranges agree with units | scale is plausible |
| derived values inspected | transformation behaved as specified |
| individuals agree with summary | result is not a summary artifact |
| uncertainty and assumptions stated | limits remain visible |

Trust points to evidence. Remaining doubt should also be recorded.

::: {.notes}
This language mirrors the final exercise section. Evidence is specific and
inspectable. “I checked carefully” is not evidence.

Ask for one remaining doubt even after every listed check passes.
:::

## AI can explain one step

Give it the local context:

```text
Here is one inspection command and its output.

1. Explain what the command checks.
2. Name one concerning result and why it matters.
3. Do not repair the data or continue the analysis.
```

Then test the explanation against the table or a small code change.

Generated code can reduce typing. The question, context, output and assumptions
still require scientific evaluation.

::: {.notes}
This matches the exercise's “Explain or debug with AI” section. The prompt is
consequential but bounded; it does not outsource the complete analysis.

Keep the original command and output beside the response. Check the explanation
with R. Use the institutionally provided Teams Copilot and do not send sensitive
data to an unapproved service.
:::

## The exercise follows the same reasoning sequence

```text
Situation
  → question and prediction
  → how the data arose
  → analysis specification
  → inspect
  → compare final weight
  → derive and compare change
  → verify
  → result
  → interpretation
  → why I trust this
```

::: {.notes}
This is a map, not a solution. Point out that the exercise deliberately asks for
both final weight and change so students can explain why the questions differ.

The analysis specification comes before the group result. AI may explain or
debug a bounded step, but the Quarto document remains the executable record.
:::

## Begin with the question

Open:

```text
exercises/tabular-comparison/analysis.qmd
```

Start with:

1. question and prediction;
2. how the data arose;
3. analysis specification;
4. inspection.

Do not begin by requesting the complete analysis from AI.

::: {.notes}
If the template is not present, run
`BIOSCI504::copy_template("lecture-3")` from the course Project. Do not
overwrite an existing exercise directory.

Give students time to read the Situation and commit to a prediction before they
see the comparison.
:::

## Questions

Then move into the `tabular-comparison` exercise.

::: {.notes}
Take questions that affect the reasoning sequence. Defer workstation-specific
troubleshooting to individual assistance during the exercise.
:::

# Further reading {.smaller}

- [Data Carpentry: Formatting Data Tables](https://datacarpentry.github.io/spreadsheet-ecology-lesson/01-format-data.html)
- [R Ecology: Exploring and Understanding Data](https://datacarpentry.github.io/R-ecology-lesson/instructor/how-r-thinks-about-data.html)
- [R Ecology: Working With Data](https://datacarpentry.github.io/R-ecology-lesson/instructor/working-with-data.html)
- [R Ecology: Data Visualization With ggplot2](https://datacarpentry.github.io/R-ecology-lesson/instructor/visualizing-ggplot.html)
- [Software Carpentry: Debugging](https://swcarpentry.github.io/python-novice-inflammation/instructor/11-debugging.html)
- [`BIOSCI504` package](https://github.com/cohmathonc/BIOSCI504)

Adapted from Data Carpentry and Software Carpentry instructional materials,
Copyright © The Carpentries, [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Changes were made for BIOSCI 504; no endorsement is implied.

::: {.notes}
The executable table sample and dictionary excerpt use the public BIOSCI504
package at the commit recorded in the deck source and README.
:::
