---
title: "Logical operations"
subtitle: "Comparisons, conditions, and filtering"
author: "Denis O'Meally"
institute: "City of Hope"
format:
  revealjs:
    theme:
      - default
      - brand
    slide-number: c/t
    transition: none
    controls: true
    progress: true
    hash: true
    history: true
    center: false
    width: 1600
    height: 900
    margin: 0.08
    code-overflow: wrap
    embed-resources: true
    footer: "BIOSCI 504 · Logical operations"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(BIOSCI504)
library(tidyverse)
data(mouse_trial)

mouse_trial <- mouse_trial |>
  as_tibble() |>
  mutate(weight_change_g = final_weight_g - baseline_weight_g)

trial_with_missing_check <- mouse_trial |>
  mutate(
    missing_required_weight = is.na(baseline_weight_g) |
      is.na(final_weight_g)
  )

complete_mouse_records <- trial_with_missing_check |>
  filter(!missing_required_weight)

mouse_conditions <- complete_mouse_records |>
  mutate(
    received_treatment = treatment == "treatment",
    gained_at_least_3_g = weight_change_g >= 3,
    meets_both_conditions = received_treatment & gained_at_least_3_g
  )
```

## By the end

You should be able to:

1. recognise that comparisons return `TRUE`, `FALSE`, or `NA`;
2. identify missing measurements and combine logical conditions;
3. use a logical condition to filter rows and inspect the selected IDs;
4. use one logical value in a simple `if`/`else` decision;
5. recognise how a working filter can become a small function; and
6. follow a supplied loop that repeats the filter.

::: {.notes}
- Use these outcomes as the roadmap for the session.
- Comparisons, a simple conditional, and filtering are the core exercise.
- The function and loop briefly reuse the same filter.
:::

## Mouse trial

### Which treated mice gained at least 3 g?

We need three columns:

- `mouse_id`
- `treatment`
- `weight_change_g`

::: {.notes}
- Resume the experiment from Day 2; one row represents one mouse.
- The 3 g threshold is an example used to make the R operations concrete.
- No treatment-effect conclusion is being made from this filter.
:::

## Weight change

### Which measurements are required?

```{r}
#| echo: true
mouse_trial |>
  select(
    mouse_id,
    treatment,
    baseline_weight_g,
    final_weight_g,
    weight_change_g
  ) |>
  slice(c(8, 12, 31, 37))
```

An unavailable baseline or final weight produces an unavailable change.

::: {.notes}
- M012 lacks a baseline measurement and M031 lacks a final measurement.
- `weight_change_g` is `NA` for both records because subtraction requires two values.
- The next step identifies these rows explicitly.
:::

## Missing values

### Which records are missing a required weight?

```{r}
#| echo: true
trial_with_missing_check <- mouse_trial |>
  mutate(
    missing_required_weight = is.na(baseline_weight_g) |
      is.na(final_weight_g)
  )

trial_with_missing_check |>
  select(
    mouse_id,
    baseline_weight_g,
    final_weight_g,
    missing_required_weight
  ) |>
  slice(c(8, 12, 31, 37))
```

`|` means logical **OR**. The result is `TRUE` when either condition is true.

::: {.notes}
- `is.na()` returns one logical value for each measurement.
- OR combines the baseline-missing and final-missing conditions.
- Source: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html
:::

## Filtering rows

### How do we keep records with both weights?

```{r}
#| echo: true
complete_mouse_records <- trial_with_missing_check |>
  filter(!missing_required_weight)

complete_mouse_records |>
  summarise(rows = n())
```

`!` means logical **NOT**. It reverses `TRUE` and `FALSE`.

`filter()` retains `TRUE` rows. Rows with `FALSE` or `NA` are not retained.

::: {.notes}
- NOT reverses the missingness condition, leaving 46 complete records.
- `filter()` retains rows for which the supplied condition is `TRUE`.
- Source: https://dplyr.tidyverse.org/reference/filter.html
:::

## Comparisons

### What does “at least 3 g” return?

```r
c(2.5, 3.0, 3.5, NA) >= 3
```

| Weight change | Result |
|---:|:---|
| 2.5 g | ? |
| 3.0 g | ? |
| 3.5 g | ? |
| unavailable | ? |

::: {.notes}
- Ask for the four results before revealing the next slide.
- Pay particular attention to exactly 3.0 g and the unavailable value.
- Keep this slide visible long enough for students to commit to an answer.
:::

## Comparisons

### What does “at least 3 g” return?

```{r}
#| echo: true
c(2.5, 3.0, 3.5, NA) >= 3
```

| Weight change | Result |
|---:|:---|
| 2.5 g | `FALSE` |
| 3.0 g | `TRUE` |
| 3.5 g | `TRUE` |
| unavailable | `NA` |

::: {.notes}
- `>=` means greater than or equal to, corresponding to “at least.”
- An unavailable value produces an unavailable comparison result.
- Source: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Comparison.html
:::

## Comparisons

### Does exactly 3.0 g meet the condition?

```{r}
#| echo: true
3 > 3
3 >= 3
3 < 3
3 <= 3
```

“At least 3.0 g” translates to `>= 3`.

::: {.notes}
- `>` means greater than; `>=` means greater than or equal to.
- `<` means less than; `<=` means less than or equal to.
- All four expressions run, but only `>= 3` matches the stated condition.
- This is the one-character error students should be able to recognize later.
:::

## Conditional execution

### Which branch will run?

```r
weight_change_g <- 3

if (weight_change_g >= 3) {
  "include"
} else {
  "exclude"
}
```

What will R return?

::: {.notes}
- Ask for a prediction before revealing the next slide.
- The comparison produces the single logical value used by `if`.
:::

## Conditional execution

### Which branch will run?

```{r}
#| echo: true
weight_change_g <- 3

if (weight_change_g >= 3) {
  "include"
} else {
  "exclude"
}
```

`if` evaluates one logical value and runs one branch.

::: {.notes}
- The condition is `TRUE`, so R returns `"include"`.
- This example makes one decision about one value.
- `filter()` applies a logical condition across table rows.
:::

## Comparing text

### Which complete mice received treatment?

```{r}
#| echo: true
mouse_conditions <- complete_mouse_records |>
  mutate(
    received_treatment = treatment == "treatment"
  )

mouse_conditions |>
  count(treatment, received_treatment)
```

`==` means **equal to**. `!=` means **not equal to**.

::: {.notes}
- The comparison is performed once for every row.
- The output is a logical column containing `TRUE` or `FALSE`.
- Character comparisons require the exact stored treatment label.
:::

## Comparing numbers

### Which complete mice gained at least 3 g?

```{r}
#| echo: true
mouse_conditions <- mouse_conditions |>
  mutate(
    gained_at_least_3_g = weight_change_g >= 3
  )

mouse_conditions |>
  count(gained_at_least_3_g)
```

The comparison is now attached to each complete mouse record.

::: {.notes}
- There are five complete records with gains of at least 3 g.
- The treatment condition has not yet been applied.
- The next step combines the two logical columns.
:::

## Combining conditions

### When is AND true?

| `mouse_id` | Received treatment? | Gained at least 3 g? | Both? |
|:---|:---:|:---:|:---:|
| M001 | `TRUE` | `FALSE` | ? |
| M008 | `TRUE` | `TRUE` | ? |
| M048 | `FALSE` | `FALSE` | ? |

```r
received_treatment & gained_at_least_3_g
```

::: {.notes}
- Ask students to fill the final column before revealing it.
- AND returns true only when the two input conditions are both true.
- M008 is the only displayed row expected to pass both conditions.
:::

## Combining conditions

### When is AND true?

| `mouse_id` | Received treatment? | Gained at least 3 g? | Both? |
|:---|:---:|:---:|:---:|
| M001 | `TRUE` | `FALSE` | `FALSE` |
| M008 | `TRUE` | `TRUE` | `TRUE` |
| M048 | `FALSE` | `FALSE` | `FALSE` |

```r
received_treatment & gained_at_least_3_g
```

`&` means logical **AND**.

::: {.notes}
- Compare each revealed result with the two input columns.
- `&` is element-wise, so each row receives its own result.
- The same operation is applied to all 46 complete records next.
:::

## Combining conditions

### How many complete records meet both?

```{r}
#| echo: true
mouse_conditions <- mouse_conditions |>
  mutate(
    meets_both_conditions = received_treatment &
      gained_at_least_3_g
  )

mouse_conditions |>
  count(meets_both_conditions)
```

Five complete records meet both conditions.

::: {.notes}
- The two component columns were inspected before they were combined.
- The combined logical column remains attached to the source records.
- Five `TRUE` values predict five rows after filtering.
:::

## Filtering rows

### Which mice meet both conditions?

```{r}
#| echo: true
treated_mice_meeting_gain_threshold <- mouse_conditions |>
  filter(meets_both_conditions) |>
  select(mouse_id, treatment, weight_change_g)

treated_mice_meeting_gain_threshold
```

The output contains five rows and five mouse IDs.

::: {.notes}
- `filter()` uses the logical column to retain the five `TRUE` rows.
- Ask students to reconcile the row count with the preceding count table.
- IDs make the selected records traceable to the source data.
:::

## Functions

### Which parts of the filter can change?

```r
mouse_data |>
  dplyr::filter(
    treatment == "treatment",
    weight_change_g >= minimum_weight_gain_g
  )
```

The data and the minimum gain are the two inputs that can change.

`dplyr::filter()` explicitly identifies the tidyverse function.

::: {.notes}
- Point to `mouse_data` and `minimum_weight_gain_g`; these become function arguments.
- The two filter conditions are the same conditions already examined.
- Avoid introducing additional function terminology on this slide.
:::

## Functions

### How do we give the filter a reusable name?

```{r}
#| echo: true
filter_treated_mice_by_weight_gain <- function(
  mouse_data,
  minimum_weight_gain_g
) {
  mouse_data |>
    dplyr::filter(
      treatment == "treatment",
      weight_change_g >= minimum_weight_gain_g
    )
}
```

::: {.notes}
- The function receives a mouse table and a minimum gain, then returns a filtered table.
- Ask students to locate the familiar filter inside the function.
- Do not read the function line by line; the new idea is giving the operation inputs and a name.
:::

## Functions

### Does it reproduce the five-row result?

```{r}
#| echo: true
filter_treated_mice_by_weight_gain(
  mouse_trial,
  minimum_weight_gain_g = 3
) |>
  select(mouse_id, treatment, weight_change_g)
```

The call reproduces the five-row result.

::: {.notes}
- Connect the two supplied arguments with the two function inputs.
- The returned table should match the earlier direct filter.
- This agreement is the only function check needed here.
:::

## Iteration

### What happens when the filter is repeated?

```r
for (minimum_weight_gain_g in c(2, 3, 4)) {
  selected_mice <- filter_treated_mice_by_weight_gain(
    mouse_trial,
    minimum_weight_gain_g
  )

  print(nrow(selected_mice))
}
```

How many times will the body run? What should happen to the row count?

::: {.notes}
- The loop body runs three times, once for each value in the vector.
- Ask whether a larger minimum gain should select more or fewer mice.
- Students are expected to follow this supplied pattern, not reproduce it from memory.
:::

## Iteration

### What happens when the filter is repeated?

```{r}
#| echo: true
for (minimum_weight_gain_g in c(2, 3, 4)) {
  selected_mice <- filter_treated_mice_by_weight_gain(
    mouse_trial,
    minimum_weight_gain_g
  )

  print(nrow(selected_mice))
}
```

::: {.notes}
- The observed counts are 21, 5, and 1 for minimum gains of 2, 3, and 4 g.
- Increasing the minimum gain makes the filter more restrictive for these data.
- The loop repeats one known operation; no new analytical rule is introduced.
:::

## Hands-on exercise

### How do you open the exercise?

In the BIOSCI504 course Project, run:

```r
BIOSCI504::copy_template("explicit-analytical-rules")
```

Then open:

```text
exercises/explicit-analytical-rules/analysis.qmd
```

::: {.notes}
- If the directory already exists, students should open it without overwriting their work.
- The exercise will follow the same comparison, combination, and filtering sequence.
- The TAs will circulate during the hands-on work.
:::

## Hands-on exercise

### What will you do?

You will:

1. inspect the relevant mouse-trial columns;
2. create and inspect logical conditions;
3. filter the treated mice meeting the gain threshold;
4. complete the small filtering function; and
5. run the supplied loop at three thresholds.

Restart R or RStudio, render again, and inspect `analysis.html`.

::: {.notes}
- The core task is the logical filter; the function and loop reuse it with support.
- Students should keep `mouse_id` beside intermediate and final results.
- A clean render confirms that the document contains the objects it uses.
:::

## Questions

::: {.notes}
- Address questions about operators, logical values, or filtering before the exercise begins.
- Function or loop syntax can be revisited with the TAs during the supported exercise.
:::

# Further reading {.smaller}

- [R comparisons](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Comparison.html)
- [R logical operators](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html)
- [`is.na()` and missing values](https://stat.ethz.ch/R-manual/R-devel/library/base/html/NA.html)
- [`dplyr::filter()`](https://dplyr.tidyverse.org/reference/filter.html)
- [Software Carpentry: Functions](https://swcarpentry.github.io/r-novice-gapminder/instructor/10-functions.html)
- [Software Carpentry: Repeating operations](https://swcarpentry.github.io/r-novice-gapminder/instructor/07-control-flow.html#repeating-operations)
- [`BIOSCI504` package](https://github.com/cohmathonc/BIOSCI504)
